---
title:  |
        | Chapter 4
        | Descriptive Analysis of Network Graph Characteristics
subtitle: "Statistical Analysis of Network Data, with R - Eric D. Kolaczyk"
author: "Thu Nguyen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true # table of content true
    toc_depth: 2  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
  html_document:
    df_print: paged
number_sections: yes
geometry: margin=2cm
---
<style type="text/css">
  .main-container {
  max-width: 800px !important;
  font-size: 18px;
  }
  code.r{
    font-size: 18px;
  }
  pre {
    font-size: 18px
  }
  h1.title {
    font-size: 30px;
    color: red;
  }
  h1 {
    font-size: 24px;
    color: blue;
  }
  h2 {
    font-size: 18px;
    color: blue;
  }
  h3 {
    font-size: 12px;
  }
</style>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen=999)
par(mar=c(0,0,1,0))
```

Libraries
```{r}
library(igraph)
library(igraphdata)
library(sand)
library(ape)
```


# Introduction

***

\clearpage

# Vertex and Edge Characteristics

***

## Vertex degree

Given $G=(V,E)$, with degree $d_v$ for vertex $v$, define $f_d$ to be the fraction of vertices $v \in V$ with degree $d_v = d$. Then $\{f_d\}_{d \geq 0}$  is the \textbf{degree distribution} of $G$. For example:

```{r fig.height=2.5,fig.width=6}
data(karate)
hist(degree(karate), col = 'lightblue', xlim = c(0, 50), 
     xlab = 'Vertex Degree', ylab = 'Frequency', main = 'Degree distribution on karate data')
```

For a \textit{weighted networks}, \textbf{vertex strength} is the sum of weights of edges incident to a given vertex, by function \texttt{\color{blue}{graph.strength()}}. For example:

```{r fig.height=2.5,fig.width=6}
hist(graph.strength(karate), col = "pink",
     xlab = "Vertex Strength", ylab = "Frequency", main = 'Vertex Strength')
```

Another dataset, `yeast` from package `igraphdata`:

```{r}
data(yeast)
nv <- vcount(yeast); ne <- ecount(yeast)
```

where there are $`r nv`$ vertices and $`r ne`$ edges, and the distributions of $degree$ and $\log(degree)$, which gives a fairly linear relationship:

```{r fig.height=2,fig.width=8}
par(mfrow = c(1,2)); par(mar=c(4,5,1.5,5))
d.yeast <- degree(yeast)
hist(d.yeast, col = 'blue', xlab = 'Vertex Degree', ylab = 'Frequency', 
     main = 'Degree Distribution')
dd.yeast <- degree.distribution(yeast)
d <- 1:max(d.yeast) - 1
ind <- (dd.yeast != 0)
plot(d[ind], dd.yeast[ind], log = 'xy', col = 'blue',
     xlab = 'Log-Degree', ylab = 'Log-Intensity', main = 'Log-Log Degree Distribution')
```

Alternatively, we can look at the average degree of the neighbors of a given vertex, by \texttt{\color{blue}{graph.knn()}}:

```{r fig.height=2,fig.width=6}
par(mar=c(4,5,.5,5))
a.nn.deg.yeast <- graph.knn(yeast,V(yeast))$knn
plot(d.yeast, a.nn.deg.yeast, log = "xy", col = "goldenrod", 
     xlab = 'Log Vertex Degree', ylab ='Log Average')
```

***

\clearpage

## Vertex Centrality

\textbf{Closeness centrality measures} attempts to measure that a vertex is 'central'   if it is 'close' to many other vertices. Let $dist(v,u)$ be the geodesic distance, the standard approach is:
$$ c_{Cl} \ (v) = \frac{1}{\sum_{u \in V} dist(v,u)} $$

\textbf{Betweenness centrality measures} attempts to summarize the extent to which a vertex is located 'between' other pairs of vertices. The perspective is that 'importance' relates to where a vertex is located relative to network' paths. Let $\sigma(s,t|v)$ be the total number of shortest paths between $s$ and $t$ passing through $v$, and $sigma(s,t)$  be the total number of shortest paths anywhere, the standard approach is:
$$ c_B \ (v) = \sum_{s \neq t \neq t \in V} \ \frac{\sigma(s,t|v)}{\sigma(s,t)} $$

\textbf{Status/Prestige/Rank}: the more central a vertex's neighbors are, the more central that vertex itself is. This is typically expressed as eigenvector solutions of linear systems of equations, called \textbf{eigenvector centrality measures}. Let $\textbf{c}_{Ei} = (c_{Ei}(1), \dots, c_{Ei}(N_v))^T$ be the solution to $\textbf{Ac}_{Ei} = \alpha^{-1}\textbf{c}_{Ei}$, where $\textbf{A}$ is the adjacency matrix:
$$ c_{Ei} \ (v) = \alpha \sum_{\{u,v\} \in E} \ c_{Ei} \ (u) $$

To display vertex centrality, use \texttt{\color{blue}{gplot.target()}} from package `sna`, to get adjacency matrix: \texttt{\color{blue}{get.adjacency()}}:

Similarly, instead of `degree(g)`, we can plot with attributes \texttt{\color{blue}{closeness(g), betweenness(g), evcent(g)\$vector}} for the 3 measures above respectively.

```{r fig.height=4,fig.width=8}
library(network); library(sna)
par(mfrow = c(2,2)); par(mar=c(0,0,1,0))
A <- get.adjacency(karate, sparse = FALSE)
g <- network::as.network.matrix(A)
sna::gplot.target(g, degree(g), main = 'Degree',
                  circ.lab = FALSE, circ.col = 'darkblue', usearrows = FALSE,
                  vertex.col = c('blue', rep('red', 32), 'yellow'), edge.col = 'darkgray')
sna::gplot.target(g, closeness(g), main = 'Closeness',
                  circ.lab = FALSE, circ.col = 'darkblue', usearrows = FALSE,
                  vertex.col = c('blue', rep('red', 32), 'yellow'), edge.col = 'darkgray')
sna::gplot.target(g, betweenness(g), main = 'Betweenness',
                  circ.lab = FALSE, circ.col = 'darkblue', usearrows = FALSE,
                  vertex.col = c('blue', rep('red', 32), 'yellow'), edge.col = 'darkgray')
sna::gplot.target(g, evcent(g), main = 'Eigenvalue',
                  circ.lab = FALSE, circ.col = 'darkblue', usearrows = FALSE,
                  vertex.col = c('blue', rep('red', 32), 'yellow'), edge.col = 'darkgray')
```

Extension from \textit{undirected} to \textit{directed} graphs are straightforward. For example: new dataset AIDS blog network:

```{r fig.height=3,fig.width=8}
l <- layout.kamada.kawai(aidsblog)
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
plot(aidsblog, layout = l, main = 'Hubs', vertex.label = '',
     vertex.size = 10*sqrt(hub.score(aidsblog)$vector))
plot(aidsblog, layout = l, main = 'Authorities', vertex.label = '',
     vertex.size = 10*sqrt(authority.score(aidsblog)$vector))

```

***

\clearpage

## Characterizing Edges

```{r}
eb <- edge.betweenness(karate)
E(karate)[order(eb, decreasing = T)[1:3]]
```

***

\clearpage

# Characterizing Network Cohesion

Questions to consider regarding \textbf{network cohesion}:

*  Do friends of a given person in a social network tend to be friends of another as well?
*  Does the structure of WWW pages tend to separate with respect to distinct types of content?

***

## Subgraphs and Censuses

\textbf{Cliques} are complete subgraphs, and thus subsets of fully cohesive vertices, like $K_1, K_2, K_3, \dots$, by \texttt{\color{blue}{cliques()}}. Example from `karate` where there are 34 isolated vertices ($K_1$), 78 pairs/$K_2$, 45 triangles/$K_3$, and so on:

```{r}
table(sapply(cliques(karate), length))
```

The 2 biggest cliques of size 5, both having the head instructor, are:
```{r}
cliques(karate)[sapply(cliques(karate), length) == 5]
```

\textbf{Maximal cliques}: cliques that are not subsets of a larger clique, \texttt{\color{blue}{maximal.cliques()}}
```{r}
table(sapply(maximal.cliques(karate), length))
```

\clearpage

\textbf{$k$-core} of $G$ is \textit{maximal} subgraph for which all vertex degrees are $\geq k$. The notion of cores is popular in visualization in helping decomposing a network into 'layers', \texttt{\color{blue}{graph.coreness()}}:

```{r fig.height=3,fig.width=6}
library(sna)
cores <- graph.coreness(karate)
par(mar=c(0,0,1,0))
sna::gplot.target(g, cores, circ.lab = FALSE, circ.col = 'darkblue', usearrows = FALSE,
                  vertex.col = cores, edge.col = 'darkgray', main = 'k-core from karate')
detach('package:sna'); detach('package:network')
```

***

\clearpage

## Density and Related Notions of Relative Frequency

\textbf{Density}: frequency of realized edges relative to potential edges. For a \textit{simple, undirected} $G$, with $n$ vertices and $m$ edges, and subgraph $H$:
$$ den(H) = \frac{m}{\frac{n(n-2)}{2}} $$
For a \textit{simple, directed} $G$:
$$ den(H) = \frac{m}{n(n-1)} $$

If $H=G$, $den(H) = den(G)$. If $H = H_v: V(H_v) = N(v)$ (set of neighbors of $v$), $den(H)$ measures the density in the immediate neighborhood of $v$. To get $N(v)$: \texttt{\color{blue}{neighborhood()}}, subgraph: \texttt{\color{blue}{induced.subgraph()}}, and density: \texttt{\color{blue}{graph.density()}}. For example:

```{r}
# Getting subgraph of neighbors of the instructor, vertex 1, and admin, vertex 34
ego.instr <- induced.subgraph(karate, neighborhood(karate, 1, 1)[[1]])
ego.admin <- induced.subgraph(karate, neighborhood(karate, 1, 34)[[1]])
den.k <- graph.density(karate); 
den.i <- graph.density(ego.instr); den.a <- graph.density(ego.admin)
```

Whereas the whole network density is low, at `r den.k`, densities at both the instructor = `r den.i` and the admin = `r den.a` are quite higher, which is consistent with the disparity in the number of within-versus between-fraction edges.

***

\textbf{Clustering coefficient}, given $\tau_{\Delta}(G)$: the number of triangles in $G$, and $\tau_3(G)$: the number of connected triples (or 2-star):
$$ cl_T(G) = \frac{3\ \tau_{\Delta}(G)}{\tau_3(G)} $$

$c|_T$ also \textbf{transitivity}, or 'fraction of transitive triples', measuring global clustering for the entire network, \texttt{\color{blue}{transitivity()}}:

```{r}
transitivity(karate)
```

\textbf{\textit{Locally}}, let $\tau_{\Delta}(v)$: the number of triangles in $G$ containing $v$, and $\tau_3(v) = \binom{d_v}{2}$: the number of connected triples for which both 2 edges are incident to $v$. Then for $v$ such that $\tau_3(v) > 0$, the \textbf{local clustering coefficient} is:
$$ cl(v) = \frac{\tau_{\Delta}(v)}{\tau_3(v)} $$

```{r}
transitivity(karate, 'local', vids = c(1,34))
```

Unique to \textit{directed} graph is \textbf{reciprocity}. There are 2 approaches: through dyads or directed edges:

```{r}
def <- reciprocity(aidsblog, mode = 'default'); rat <- reciprocity(aidsblog, mode = 'ratio')
print(cat('Dyads: ', def, '\n', 'Directed edges: ', rat, '\n'))
```

***

## Connectivity, Cuts, and Flows

Recall \textit{components}, by \texttt{\color{blue}{decompse.graph()}}, and getting vertex count by \texttt{\color{blue}{vcount()}}. From `yeast` dataset, there is \textit{giant component} of $2375$ vertices or $90\%$ of all vertices, $0$ isolated vertices, $63$ pairs, ...

```{r}
comps <- decompose.graph(yeast)
table(sapply(comps, vcount))
```

```{r}
yeast.gc <- decompose.graph(yeast)[[1]]  # giant component
ave.path <- round(average.path.length(yeast.gc),2)
diam <- diameter(yeast.gc)
tran <- round(transitivity(yeast.gc),2)
```

The giant component have a few characteristics of a \textbf{\textit{small world}} model such as small \textit{shortest-path distance between pairs of vertices} of `r ave.path`, and small \textit{longest paths} of `r diam`, and high \textit{clustering} of `r tran`, indicating that close to $50 \%$ of connected triples form triangles.

***

Recall from Graph Theory of Math 154, \textbf{$k$-vertex-connected} and \textbf{$k$-connected}, and that $\kappa(u,v) \leq \lambda(u,v) \leq \delta(G)$. To see connectivity: \texttt{\color{blue}{vertex.connectivity()}}, and \texttt{\color{blue}{edge.connectivity()}}:

```{r}
v.con <- vertex.connectivity(yeast.gc)
e.con <- edge.connectivity(yeast.gc)
print(paste('Vertex connectivity: ', v.con, '; Edge connectivity: ', e.con))
```

\textbf{Vertex-cut (edge-cut)} is a set vertices (edges) that disconnect $G$. Also, if only of size 1, \textit{cut vertex} or \textit{articulation point}, by \texttt{\color{blue}{articulation.points()}}:

```{r}
yeast.cut.vertices <- articulation.points(yeast.gc)
print(paste('The number of cut vertices:', length(yeast.cut.vertices)))
```

\textbf{Menger's theorem} vertex form:
$$ \min\{ab-\text{separator} \} = \max\{\text{pariwise internally disjoint paths between $a$ and $b$}\} $$

`R`: \texttt{\color{blue}{shortest.paths()}}, \texttt{\color{blue}{graph.maxflow()}}, and \texttt{\color{blue}{graph.mincut()}}. 

\clearpage

These concepts extend naturally to \textit{directed} graphs:

```{r}
w.con <- is.connected(aidsblog, mode = 'weak')
s.con <- is.connected(aidsblog, mode = 'strong')
print(paste0('Weak: ', w.con, '; Strong: ', s.con))
```

Strongly connected components, from \texttt{\color{blue}{clusters()}}:

```{r}
aidsblog.scc <- clusters(aidsblog, mode = 'strong')
table(aidsblog.scc$csize)
```

***

\clearpage

# Graph Partitioning

***

## Hierarchical Clustering

There are 2 main approaches:

*  \textbf{agglomerative}: successive coarsening of paritions through merging
*  \textbf{divisive}: successive coarsening of paritions through splitting

To do clustering: \texttt{\color{blue}{fastgreedy.community()}}, \texttt{\color{blue}{sizes()}}, \texttt{\color{blue}{membership()}}:

```{r}
kc <- fastgreedy.community(karate)
sizes(kc)
```

```{r}
membership(kc)
```

\texttt{\color{blue}{dendPlot()}} from package `ape` for dendogram:

```{r fig.height=8,fig.width=8}
library(ape)
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
plot(kc, karate, main = 'Clusters')
dendPlot(kc, mode = 'phylo', main = 'Dendogram')
```

***

\clearpage

## Spectral Partitioning

\textit{Spectral graph theory}: connectivity of $G$ is associated with the eigen-analysis of certain matrices. Define \textit{Laplacian} $L$ of $G$ with adjacency matrix $A$, and diagonal degree matrix $D = diag[(d_v)]$:
$$L = D - A$$

Then the number of components in $G$ is directly related to the number of non-zero eigenvalues of $L$. In `R`: \texttt{\color{blue}{graph.laplacian()}}, \texttt{\color{blue}{eigen()}}, \texttt{\color{blue}{get.vertex.attribute()}}:

```{r fig.height=1.5,fig.width=6}
par(mar=c(2,5,.5,5))
k.lap <- graph.laplacian(karate)
eig.anal <- eigen(k.lap)
plot(eig.anal$values, col = 'blue', ylab = 'Eigenvalues')
```

Extracting \textit{Fiedler vector} $\textbf{x}_2$ from:
$$ S = \{ v \in V: \textbf{x}_2(v) \geq 0 \}, \ \ \text{ and } \ \ 
  \bar{S} = \{ v \in V: \textbf{x}_2(v) < 0 \}$$

which partition into 2 subsets of vertices, for example:

```{r fig.height=2,fig.width=6}
par(mar=c(4,5,.5,5))
f.vec <- eig.anal$vectors[, 33]
faction <- get.vertex.attribute(karate, 'Faction')
f.colors <- as.character(length(faction))
f.colors[faction == 1] <- 'red'; f.colors[faction == 2] <- 'cyan'
plot(f.vec, pch = 16, xlab = 'Actor Number', ylab = 'Fiedler Vector', col = f.colors)
abline(0, 0, lwd = 2, col = 'lightgray')
```

***

\clearpage

## Validation of Graph Partitioning

***

# Assortativity and Mixing

```{r}
assortativity.degree(yeast)
```

***