---
title:  |
        | Chapter 3
        | Visualizing Network Data
subtitle: "Statistical Analysis of Network Data, with R - Eric D. Kolaczyk"
author: "Thu Nguyen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true # table of content true
    toc_depth: 2  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
  html_document:
    df_print: paged
number_sections: yes
geometry: margin=2cm
---
<style type="text/css">
  .main-container {
  max-width: 800px !important;
  font-size: 18px;
  }
  code.r{
    font-size: 18px;
  }
  pre {
    font-size: 18px
  }
  h1.title {
    font-size: 30px;
    color: red;
  }
  h1 {
    font-size: 24px;
    color: blue;
  }
  h2 {
    font-size: 18px;
    color: blue;
  }
  h3 {
    font-size: 12px;
  }
</style>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen=999)
par(mar=c(0,0,1,0))
```

Libraries
```{r}
library(igraph)
library(igraphdata)
library(sand)
```

***

# Introduction

***

# Elements of Graph Visualization

***

\clearpage

# Graph Layouts

```{r}
g.l <- graph.lattice(c(5,5,5)) # Lattice graph
# data(aidsblog)
summary(aidsblog)
```

\textit{Circular layout}: \texttt{\color{blue}{layout=layout.circle}}:

```{r fig.height=1.5,fig.width=8,fig.align='center'}
igraph.options(vertex.size=3, vertex.label=NA, edge.arrow.size=0.2)
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
plot(g.l, layout=layout.circle); title("5x5x5 Lattice")
plot(aidsblog, layout=layout.circle); title("Blog Network")
```

\textit{Spring-Embedder layout}: \texttt{\color{blue}{layout=layout.fruchterman.reingold}}:

```{r fig.height=1.5,fig.width=8,fig.align='center'}
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
plot(g.l, layout=layout.fruchterman.reingold); title("5x5x5 Lattice")
plot(aidsblog, layout=layout.fruchterman.reingold); title("Blog Network")
```

\textit{Energy-Placement layout}: \texttt{\color{blue}{layout=layout.kamada.kawai}}:

```{r fig.height=1.5,fig.width=8,fig.align='center'}
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
plot(g.l, layout=layout.kamada.kawai); title("5x5x5 Lattice")
plot(aidsblog, layout=layout.kamada.kawai); title("Blog Network")
```

\textbf{Trees}: 

*  \textit{circular}: \texttt{\color{blue}{layout=layout.circle}}
*  \textit{radial}: \texttt{\color{blue}{layout=layout.reingold.tilford(circular=T)}}
*  \textit{layered}: \texttt{\color{blue}{layout=layout.reingold.tilford}}

```{r fig.height=1.5,fig.width=8,fig.align='center'}
g.tree <- graph.formula(1-+2,1-+3,1-+4,2-+5,2-+6,2-+7,3-+8,3-+9,4-+10)
par(mfrow=c(1, 3)); par(mar=c(0,0,1,0))
igraph.options(vertex.size=30, edge.arrow.size=0.5, vertex.label=NULL)
plot(g.tree, layout=layout.circle); title("Circular")
plot(g.tree, layout=layout.reingold.tilford(g.tree, circular=T)); title("Radial")
plot(g.tree, layout=layout.reingold.tilford); title("Layered")
```

\textbf{Bipartite} graphs: \texttt{\color{blue}{layout=layout.bipartite}}

```{r fig.height=1.5,fig.width=8,fig.align='center'}
par(mar=c(0,0,0,0))
plot(g.bip, layout=-layout.bipartite(g.bip)[,2:1],
     vertex.size=30, vertex.shape=ifelse(V(g.bip)$type, "rectangle", "circle"),
     vertex.color=ifelse(V(g.bip)$type, "red", "cyan"))
```

***

\clearpage

# Decorating Graph Layouts

`Karate` dataset

```{r fig.height=3,fig.width=8,fig.align='center'}
library(igraphdata)
data(karate)
# Reproducible layout
set.seed(42)
l <- layout.kamada.kawai(karate)
# Plot undecorated first.
igraph.options(vertex.size=10)
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
plot(karate, layout=l, vertex.label=V(karate)); title('Undecorated')
# Now decorate, starting with labels.
V(karate)$label <- sub("Actor ", "", V(karate)$name)
# Two leaders get shapes different from club members.
V(karate)$shape <- "circle"
V(karate)[c("Mr Hi", "John A")]$shape <- "rectangle"
# Differentiate two factions by color.
V(karate)[Faction == 1]$color <- "red"
V(karate)[Faction == 2]$color <- "dodgerblue"
# Vertex area proportional to vertex strength
# (i.e., total weight of incident edges).
V(karate)$size <- 4*sqrt(graph.strength(karate))
V(karate)$size2 <- V(karate)$size * .5
# Weight edges by number of common activities
E(karate)$width <- E(karate)$weight
# Color edges by within/between faction.
F1 <- V(karate)[Faction==1]
F2 <- V(karate)[Faction==2]
E(karate)[ F1 %--% F1 ]$color <- "pink"
E(karate)[ F2 %--% F2 ]$color <- "lightblue"
E(karate)[ F1 %--% F2 ]$color <- "yellow"
# Offset vertex labels for smaller points (default=0).
V(karate)$label.dist <- ifelse(V(karate)$size >= 10, 0, 0.75)
# Plot decorated graph, using same layout.
plot(karate, layout=l); title('Decorated')
```

\clearpage

`Lazega` dataset

```{r fig.height=3,fig.width=8,fig.align='center'}
library(sand)
data(lazega)
par(mar=c(0,0,1,0))
# Office location indicated by color.
colbar <- c("red", "dodgerblue", "goldenrod")
v.colors <- colbar[V(lazega)$Office]
# Type of practice indicated by vertex shape.
v.shapes <- c("circle", "square")[V(lazega)$Practice]
# Vertex size proportional to years with firm.
v.size <- 3.5*sqrt(V(lazega)$Years)
# Label vertices according to seniority.
v.label <- V(lazega)$Seniority
# Reproducible layout.
set.seed(42)
l <- layout.fruchterman.reingold(lazega)
plot(lazega, layout=l, vertex.color=v.colors,
     vertex.shape=v.shapes, vertex.size=v.size,
     vertex.label=v.label)
```

***

\clearpage

# Visualizing Large Networks

`fblog` dataset

```{r}
# summary(fblog)
party.names <- sort(unique(V(fblog)$PolParty))
party.names
```

```{r fig.height=2,fig.width=8,fig.align='center'}
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
# Layout: Kamada-Kawai
set.seed(42)
l = layout.kamada.kawai(fblog)
party.nums.f <- as.factor(V(fblog)$PolParty)
party.nums <- as.numeric(party.nums.f)
plot(fblog, layout=l, vertex.label=NA, vertex.color=party.nums, vertex.size=3)
title('Kamada-Kawai')
# Layout: DrL
set.seed(42)
l <- layout.drl(fblog)
plot(fblog, layout=l, vertex.size=5, vertex.label=NA, vertex.color=party.nums)
title('DrL')
```

\textit{Contracting group of vertices into meta-vertices:}

```{r fig.height=2,fig.width=8,fig.align='center'}
par(mar=c(0,0,0,0))
fblog.c <- contract.vertices(fblog, party.nums)
E(fblog.c)$weight <- 1
fblog.c <- simplify(fblog.c)
party.size <- as.vector(table(V(fblog)$PolParty))
plot(fblog.c, vertex.size=5*sqrt(party.size),
     vertex.label=party.names,
     vertex.color=V(fblog.c),
     edge.width=sqrt(E(fblog.c)$weight),
     vertex.label.dist=1.5, edge.arrow.size=0)
```

***

Further look into `karate` dataset: (1) Extracting neighbors:

```{r}
data(karate)
k.nbhds <- graph.neighborhood(karate, order=1)
sapply(k.nbhds, vcount)
```

(2) Pulling out 2 largest subnetworks and plotting them

```{r fig.height=2,fig.width=8,fig.align='center'}
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
k.1 <- k.nbhds[[1]]
k.34 <- k.nbhds[[34]]
plot(k.1, vertex.label=NA,
     vertex.color=c("red", rep("lightblue", 16)))
plot(k.34, vertex.label=NA,
     vertex.color=c(rep("lightblue", 17), "red"))
```

