---
title:  |
        | Chapter 5
        | Mathematical Models for Network Graphs
subtitle: "Statistical Analysis of Network Data, with R - Eric D. Kolaczyk"
author: "Thu Nguyen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true # table of content true
    toc_depth: 2  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
  html_document:
    df_print: paged
number_sections: yes
geometry: margin=2cm
---
<style type="text/css">
  .main-container {
  max-width: 800px !important;
  font-size: 18px;
  }
  code.r{
    font-size: 18px;
  }
  pre {
    font-size: 18px
  }
  h1.title {
    font-size: 30px;
    color: red;
  }
  h1 {
    font-size: 24px;
    color: blue;
  }
  h2 {
    font-size: 18px;
    color: blue;
  }
  h3 {
    font-size: 12px;
  }
</style>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen=999)
par(mar=c(0,0,1,0))
```

Libraries
```{r}
library(igraph)
library(igraphdata)
library(sand)
```

***

# Introduction

***

\clearpage

# Classical Random Graph Models

\textbf{\textit{Random graph model}} refers to model specifying a collection $\mathcal{G}$ and a \textbf{uniform probability} $\mathbb{P}$ over $\mathcal{G}$.

The classical example is \textbf{Erdos-Renyi} model: given graph $G = (V,E)$ with $n$ vertices and $m$ edges, with $N = \binom{n}{2}$:
$$ \mathbb{P} \ (G) = \binom{N}{m}^{-1} $$

A variant of $\mathcal{G}_{n,m}$ is a collection $\mathcal{G}_{n,p}$ with fixed $n$ vertices, and probability $p \in (0,1)$ for every edge between each pair of vertices \textit{independently}. The model's degree distribution is approximately $Poisson$. In `R`: \texttt{\color{blue}{erdos.renyi.game()}}:

```{r fig.height=3,fig.width=8,fig.align='center'}
par(mfrow=c(1,2)); par(mar=c(2,0,1,0))
set.seed(42)
g.er <- erdos.renyi.game(100, .02)
plot(g.er, layout = layout.circle, vertex.label = NA, main = 'Erdos-Renyi Random graph')
hist(degree(g.er), col = 'lightblue', xlab = 'Degree', ylab = 'Frequency',
     main = expression(Pois: mu: 1.9))
```

Further look: $G$ is not connected, with $15$ isolated vertices and a giant component of $71$ vertices:

```{r}
table(sapply(decompose.graph(g.er), vcount))
```

```{r}
apl <- round(average.path.length(g.er),2)
diam <- diameter(g.er)
trans <- round(transitivity(g.er),2)
print(paste0('Average path length: ', apl, '; Diameter: ', diam, '; Transitivity: ', trans))
```

***

\clearpage

# Generalized Random Graph Models

The \textbf{Erdos-Renyi} model can be generalized into a collection $\mathcal{G}$ with fixed $n$ vertices and a given characteristics, and a \textbf{uniform distribution} for all graph $G$ over $\mathcal{G}$. While \textbf{Erdos-Renyi} fixes $m$ edges, the $2^{nd}$ characteristic is an \textit{ordered} pre-specified degree sequence $\{d_1, d_2, \dots, d_n\}$. In `R`: \texttt{\color{blue}{degree.sequence.game()}}:

```{r fig.height=2,fig.width=8,fig.align='center'}
par(mfrow=c(1,2)); par(mar=c(0,0,0,0))
degs <- c(2,2,2,2,3,3,3,3)
g1 <- degree.sequence.game(degs, method = 'vl'); g2 <- degree.sequence.game(degs, method = 'vl')
plot(g1, vertex.label = NA); plot(g2, vertex.label = NA)
```

```{r}
iso <- graph.isomorphic(g1, g2); print(paste0('Are they isomorphic? ', iso))
```

```{r}
print(paste0('Edge counts: graph 1: ', ecount(g1), '; graph 2: ', ecount(g2)))
```

For example, taking the sequence from the `yeast` dataset:

```{r}
data(yeast)
degs <- degree(yeast)
fake.yeast <- degree.sequence.game(degs, method = 'vl')
print(paste0('Same degree sequence? ', all(degree(yeast) == degree(fake.yeast))))
```

```{r}
diam <- round(c(diameter(yeast), diameter(fake.yeast)),0)
trans <- round(c(transitivity(yeast), transitivity(fake.yeast)),4)
data.frame(data = c('Original Yeast', 'New Yeast'),
           diameter = diam, transitivity = trans)
```

***

\clearpage

# Network Graph Models Based on Machanisms

***

## Small-World Models

\textbf{Small-World models:} high clustering but small distances between most nodes, which cannot be produced by \textit{classical random graphs}. In `R`, \texttt{\color{blue}{watts.strogatz.game()}}:

```{r fig.height=3,fig.width=8,fig.align='center'}
par(mfrow=c(1,2)); par(mar=c(0,0,1,0))
g.ws <- watts.strogatz.game(1, 25, 5, .05)
g.lat100 <- watts.strogatz.game(1, 25, 5, 0)
plot(g.ws, layout = layout.circle, vertex.label = NA, main = 'Small-World Model: p = .05')
plot(g.lat100, layout = layout.circle, vertex.label = NA, main = 'Lattice graph: p = 0')
```

```{r}
apl <- round(c(average.path.length(g.ws), average.path.length(g.lat100)),2)
diam <- round(c(diameter(g.ws), diameter(g.lat100)),0)
trans <- round(c(transitivity(g.ws), transitivity(g.lat100)),4)
data.frame(data = c('Small-World Model', 'Lattice'),
           ave.path = apl, diameter = diam, transitivity = trans)
```

\clearpage

Simulation of the \textbf{Small-world models} as $p$ varies:

```{r fig.height=4,fig.width=8,fig.align='center'}
steps <- seq(-4, -.5, .1)
len <- length(steps)
cl <- numeric(len)
apl <- numeric(len)
ntrials <- 100
for (i in 1:len) {
  cltemp <- numeric(ntrials)
  apltemp <- numeric(ntrials)
  for (j in 1:ntrials) {
    g <- watts.strogatz.game(1, 1000, 10, 10^steps[i])
    cltemp[j] <- transitivity(g)
    apltemp[j] <- average.path.length(g)
  }
  cl[i] <- mean(cltemp)
  apl[i] <- mean(apltemp)
}
par(mar=c(4,5,2.5,2))
plot(steps, cl/max(cl), ylim = c(0,1), lwd = 3, type = 'l', col = 'blue',
     xlab = expression(log[10](p)), ylab = 'Clustering & Average Path Length')
lines(steps, apl/max(apl), lwd = 3, col = 'red')
```

***

\clearpage

## Preferential Attachment Models

\textbf{Preferential attachment}: models the principle 'the rich get richer', where the probability of the new vertex will be conneced to a given vertex $v$ is proportional to the existing degree of $v$:
$$ p = \frac{d_v}{\sum_{v' \in V} d_{v'}} $$

At each stage, $k$ existing vertices are connected to a new vertex preferential to those with higher degrees. After $t$ iterations, the resulting $G^t$ will have $N_v^t = n + t$ vertices and $N_e^t = m + tk$ edges. In `R`, \texttt{\color{blue}{barabasi.game()}}:

```{r fig.height=3,fig.width=8,fig.align='center'}
par(mfrow=c(1,2)); par(mar=c(2,0,1,0))
set.seed(42)
g.ba <- barabasi.game(100, directed = FALSE)
plot(g.ba, layout = layout.circle, vertex.label = NA)
hist(degree(g.ba), col = 'lightblue', xlab = 'Degree', ylab = 'Frequency', main = '')
```

**Remark:** as $t \rightarrow \infty$, the graphs $G^t$ have degree distributions that tend to a power-law form $d^{-\alpha}$, with $\alpha = 3$.

```{r}
data.frame(Attributes = c('Ave.Path.Length', 'Diameter', 'Transitivity'),
           Values = c(average.path.length(g.ba), diameter(g.ba), transitivity(g.ba)))
```

***

\clearpage

# Assessing Significance of Network Graph Characteristics

*  Question of interest: \textbf{is $\eta \ (G^{obs})$ significant?} where $\eta(\cdot)$ is some structural characteristic, under a formal reference distribution, assuming \textit{uniform distribution} of elements in $\mathcal{G}$:
$$ \mathbb{P}_{\eta, \mathcal{G}} \ (t) 
    = \frac{\# \{G \in \mathcal{G}: \eta(G) \leq t \}}{|\mathcal{G}|} $$

***

## Assessing the Number of Communities in a Network

Using \textit{Monte Carlo simulation}, with the $n$ vertices, $m$ edges, and degree sequence from `karate` dataset:

```{r fig.height=2.5,fig.width=8,fig.align='center'}
data(karate)
# Observed graph statistics
nv <- vcount(karate); ne <- ecount(karate); degs <- degree(karate)
# Fix #(vertices) and #(edges)
num.comm.rg <- numeric(ntrials)
for (i in 1:ntrials) {
  g.rg <- erdos.renyi.game(nv, ne, type = 'gnm')
  c.rg <- fastgreedy.community(g.rg)
  num.comm.rg[i] <- length(c.rg)
}
# Fix degree sequence
num.comm.grg <- numeric(ntrials)
for (i in 1:ntrials) {
  g.grg <- degree.sequence.game(degs, method = 'vl')
  c.grg <- fastgreedy.community(g.grg)
  num.comm.grg[i] <- length(c.grg)
}
par(mar=c(.5,5,.5,0))
results <- c(num.comm.rg, num.comm.grg)
idx <- c(rep(0, ntrials), rep(1, ntrials))
counts <- table(idx, results)/ntrials
barplot(counts, beside = TRUE, col = c('blue', 'red'),
        xlab = 'Number of Communities', ylab = 'Relative Frequency',
        legend = c('Fixed Size', 'Fixed Degree Sequence'))
```

***

\clearpage

## Assessing Small World Properties

Given a \textit{directed} graph $G$ with adjacency matrix \textbf{A}, and total degree $d_v^{tot}$ of vertex $v$ (in-degree + out-degree), the \textit{clustering coefficient} is:
$$ cl(v) = \frac{(\textbf{A} + \textbf{A}^T)^3_{vv}}{2\ [\ d_v^{tot} \ (d_v^{tot} - 1) - 2(\textbf{A}^2)_{vv}\ ]} $$

For example, dataset `macaque` from package `igraphdata`:

```{r}
library(igraphdata)
data(macaque)
summary(macaque)
```

```{r}
# function to calculate clustering coefficient for directed graph:
clust.coef.dir <- function(graph) {
  A <- as.matrix(get.adjacency(graph))
  S <- A + t(A)
  deg <- degree(graph, mode = 'total')
  numerator <- diag( S %*% S %*% S )
  denom <- diag(A %*% A)
  denominator <- 2 * (deg * (deg-1) - 2*denom)
  cl <- mean(numerator/denominator)
  return(cl)
}

ntrials <- 1000
nv <- vcount(macaque); ne <- ecount(macaque)
cl.rg <- numeric(ntrials); apl.rg <- numeric(ntrials)
for (i in 1:ntrials) {
  g.rg <- erdos.renyi.game(nv, ne, type = 'gnm', directed = TRUE)
  cl.rg[i] <- clust.coef.dir(g.rg)
  apl.rg[i] <- average.path.length(g.rg)
}
```

```{r}
obs.cl <- clust.coef.dir(macaque); obs.apl <- average.path.length(macaque)
p.cl <- sum(cl.rg <= obs.cl)/ntrials; p.apl <- sum(apl.rg <= obs.apl)/ntrials
data.frame(Attributes = c('Clustering coefficient', 'Average Path Length'),
           p.values = c(p.cl, p.apl))
```

**Interpretation of result:** `macaque` dataset has substantially more clustering, and longer shortest paths between vertex pairs than a random network, indicating that the small-world behavior is not clear in this dataset.

***